Inferring cosmology from gravitational waves using non-parametric detector-frame mass distribution

### 2024-10-29 src/tex/ms.tex
[showyourwork] overleaf sync

\abstract{
    The challenge of understanding the Universe's dynamics, particularly the Hubble tension, requires precise measurements of the Hubble constant.
    Building upon the existing spectral-siren method, which capitalizes on population information from gravitational-wave sources, this paper explores an alternative way to analyze the population data to obtain the cosmological parameters in $\Lambda$CDM.
    We demonstrated how non-parametric methods, which are flexible models that can be used to agnostically reconstruct arbitrary probability densities, can be incorporated into this framework and leverage the detector-frame mass distribution to infer the cosmological parameters.
    We tested our method with mock data and applied it to $70$ real events from the third gravitational-wave transient catalog of the LIGO-Virgo-KAGRA Collaboration.
}

@@

\begin{acronym}
    \acro{GW}{gravitational-wave}
    \acro{LVK}{LIGO-Virgo-KAGRA Collaboration}
    \acro{PE}{parameter estimation}
    \acro{CMB}{cosmic microwave background}
    \acro{EM}{electromagnetic}
    \acro{(H)DPGMM}{hierarchy of Dirichlet process Gaussian mixture models}
    \acro{BBH}{binary-black-hole}
\end{acronym}

@@

In this section, we apply our method to the real data from the \ac{LVK}.
We included all $70$ events labeled as BBH in Table~I of \cite{KAGRA:2021duu}, these correspond to all the events with $\mathrm{FAR_{min}} < 1\ \mathrm{yr}^{-1}$, where $\mathrm{FAR_{min}}$ is the smallest false alarm rate among all search pipelines.
These events are different from the events used in \cite{LIGOScientific:2021aug}, therefore, the results are not directly comparable.

@@

We then perform the remaining steps similar to Sec.~\ref{sec:inference_multi}.
We set the bounds for $\kappa$ to $[-10, 10]$, and the bounds for other free parameters are the same as in Table~\ref{tab:bounds}.
The result of the remapping is shown in Fig.~\ref{fig:real_result_multi}, and it is consistent with the posterior distribution reported in \cite{KAGRA:2021duu}.
\begin{figure}
    \script{plot_real_result_multi.py}
    \includegraphics[width=\linewidth]{figures/real_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the real data inference.
    }
    \label{fig:real_result_multi}
\end{figure}

The result implies that \variable{output/real_H0.txt}, with symmetric $90\%$-credible intervals around the median.
The uncertainty is much smaller than the one reported in \cite{KAGRA:2021duu} ($H_0 = 50^{+37}_{-30} \, \mathrm{km/s/Mpc}$), mainly because it only includes the uncertainty from the reconstruction, which underestimates the true uncertainty which also includes the mismatch between the population model and the observed population distribution.

@@

T.~N. and S.~R. acknowledge financial support from the German Excellence Strategy via the Heidelberg Cluster of Excellence (EXC 2181 - 390900948) STRUCTURES.
S.~R. acknowledges financial support from the European Research Council for the ERC Consolidator grant DEMOBLACK, under contract no. 770017. 
We acknowledge the use of computing facilities supported by grants from the Croucher Innovation Award from the Croucher Foundation Hong Kong.
This research made use of the bwForCluster Helix: the authors acknowledge support by the state of Baden-Württemberg through bwHPC and the German Research Foundation (DFG) through grant INST 35/1597-1 FUGG.
This research has made use of data or software obtained from the Gravitational Wave Open Science Center (gwosc.org), a service of the LIGO Scientific Collaboration, the Virgo Collaboration, and KAGRA. This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation, as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain. KAGRA is supported by Ministry of Education, Culture, Sports, Science and Technology (MEXT), Japan Society for the Promotion of Science (JSPS) in Japan; National Research Foundation (NRF) and Ministry of Science and ICT (MSIT) in Korea; Academia Sinica (AS) and National Science and Technology Council (NSTC) in Taiwan.
This paper was compiled using \textsc{showyourwork} \cite{Luger2021} to facilitate reproducibility.

>>>>

\abstract{
    The challenge of understanding the Universe's dynamics, particularly the Hubble tension, requires precise measurements of the Hubble constant.
    Building upon the existing spectral-siren method, which capitalizes on population information from gravitational-wave sources, this paper explores an alternative way to analyze the population data to obtain the cosmological parameters in $\Lambda$CDM.
    We demonstrated how non-parametric methods, which are flexible models that can be used to agnostically reconstruct arbitrary probability densities, can be incorporated into this framework and leverage the detector-frame mass distribution to infer the cosmological parameters.
    We tested our method with mock data and applied it to $70$ binary black hole mergers from the third gravitational-wave transient catalog of the LIGO-Virgo-KAGRA Collaboration.
}

@@

\begin{acronym}
    \acro{GW}{gravitational-wave}
    \acro{LVK}{LIGO-Virgo-KAGRA Collaboration}
    \acro{PE}{parameter estimation}
    \acro{CMB}{cosmic microwave background}
    \acro{EM}{electromagnetic}
    \acro{(H)DPGMM}{hierarchy of Dirichlet process Gaussian mixture models}
    \acro{BBH}{binary-black-hole}
    \acro{FAR}{false alarm rate}
\end{acronym}

@@

In this section, we apply our method to the \ac{GW} events released by the \ac{LVK} as part of the GWTC-2.1 \cite{LIGOScientific:2021usb} and GWTC-3 \cite{KAGRA:2021vkt} catalogs.
We included all $70$ events labeled as BBH in Table~I of \cite{KAGRA:2021duu}, these correspond to all the events with $\mathrm{FAR_{min}} < 1\, \mathrm{yr}^{-1}$, where $\mathrm{FAR_{min}}$ is the smallest \ac{FAR} among all search pipelines.
These events are different from the events used in \cite{LIGOScientific:2021aug}, where the authors use a \ac{FAR} threshold of $\mathrm{FAR_{min}} < 0.25\, \mathrm{yr}^{-1}$, therefore the results we present here might be slightly different from the ones in \cite{LIGOScientific:2021aug}.
We opted for a larger \ac{FAR} threshold -- consistent, however, with the one used in population studies such as \cite{KAGRA:2021duu} -- to have a larger number of events to include in the non-parametric analysis.

@@

We then perform the remaining steps similar to Sec.~\ref{sec:inference_multi}.
We set the bounds for $\kappa$ to $[-10, 10]$, and the bounds for other free parameters are the same as in Table~\ref{tab:bounds}.
The result of the remapping is shown in Fig.~\ref{fig:real_result_multi}, and it is consistent with the posterior distribution reported in \cite{KAGRA:2021duu}.
\begin{figure}
    \script{plot_real_result_multi.py}
    \includegraphics[width=\linewidth]{figures/real_result_multi.pdf}
    \caption{
        Optimized parameter distribution for the real data inference.
    }
    \label{fig:real_result_multi}
\end{figure}
The result implies that \variable{output/real_H0.txt}, with symmetric $90\%$-credible intervals around the median.
The uncertainty is much smaller than the one reported in \cite{KAGRA:2021duu} ($H_0 = 50^{+37}_{-30} \, \mathrm{km/s/Mpc}$), mainly because it only includes the uncertainty from the reconstruction, which underestimates the true uncertainty which also includes the mismatch between the population model and the observed population distribution.
Figure \ref{fig:real_result_multi} shows a peak in the marginalized posterior distribution of $H_0$ around $70\, \mathrm{km/s/Mpc}$.
This means a significant fraction of the reconstructed observed population distribution has an optimized $H_0$ around $70\, \mathrm{km/s/Mpc}$, which again does not include the uncertainty from the mismatch between the population model and the observed population distribution.
The peak should be interpreted as an artifact of the optimization procedure.

@@

We thank Grégoire Pierra for helpful discussions.
T.~N. and S.~R. acknowledge financial support from the German Excellence Strategy via the Heidelberg Cluster of Excellence (EXC 2181 - 390900948) STRUCTURES.
S.~R. acknowledges financial support from the European Research Council for the ERC Consolidator grant DEMOBLACK, under contract no. 770017. 
We acknowledge the use of computing facilities supported by grants from the Croucher Innovation Award from the Croucher Foundation Hong Kong.
This research made use of the bwForCluster Helix: the authors acknowledge support by the state of Baden-Württemberg through bwHPC and the German Research Foundation (DFG) through grant INST 35/1597-1 FUGG.
This research has made use of data or software obtained from the Gravitational Wave Open Science Center (gwosc.org), a service of the LIGO Scientific Collaboration, the Virgo Collaboration, and KAGRA. This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation, as well as the Science and Technology Facilities Council (STFC) of the United Kingdom, the Max-Planck-Society (MPS), and the State of Niedersachsen/Germany for support of the construction of Advanced LIGO and construction and operation of the GEO600 detector. Additional support for Advanced LIGO was provided by the Australian Research Council. Virgo is funded, through the European Gravitational Observatory (EGO), by the French Centre National de Recherche Scientifique (CNRS), the Italian Istituto Nazionale di Fisica Nucleare (INFN) and the Dutch Nikhef, with contributions by institutions from Belgium, Germany, Greece, Hungary, Ireland, Japan, Monaco, Poland, Portugal, Spain. KAGRA is supported by Ministry of Education, Culture, Sports, Science and Technology (MEXT), Japan Society for the Promotion of Science (JSPS) in Japan; National Research Foundation (NRF) and Ministry of Science and ICT (MSIT) in Korea; Academia Sinica (AS) and National Science and Technology Council (NSTC) in Taiwan.
This paper was compiled using \textsc{showyourwork} \cite{Luger2021} to facilitate reproducibility.
